---
title: "Linear Regression Model, Predictions, Accuracy"
author: "Devan Morehouse"
date: "2023-09-20"
output: html_document
---

# Linear Regression Model

Loading the data
```{r}
car.df <- read.csv("C:\\Users\\dev46\\OneDrive\\Desktop\\School Documents\\Spring 2023\\MGQ 408 Bus. Analytics & Data Science\\Data\\ToyotaCorolla.csv")

# use first 1000 rows of data
car.df <- car.df[1:1000, ]
# select variables for regression
selected.var <- c(3, 4, 7, 8, 9, 10, 12, 13, 14, 17, 18)
```

## Partition data

```{r}
set.seed(1)  # set seed for reproducing the partition
train.index <- sample(c(1:1000), 600)  
train.df <- car.df[train.index, selected.var]
valid.df <- car.df[-train.index, selected.var]
```

## Training the model

```{r}
# use lm() to run a linear regression of Price on all 11 predictors in the
# training set. 
# use . after ~ to include all the remaining columns in train.df as predictors.
car.lm <- lm(Price ~ ., data = train.df)

#  use options() to ensure numbers are not displayed in scientific notation.
options(scipen = 999)
summary(car.lm)
```

## Making predictions on a new set and testing the accuracy

```{r}
library(forecast)
car.lm.pred <- predict(car.lm, valid.df)
some.residuals <- valid.df$Price[1:20] - car.lm.pred[1:20]
data.frame("Predicted" = car.lm.pred[1:20], "Actual" = valid.df$Price[1:20],
           "Residual" = some.residuals)

options(scipen=999, digits = 3)
# use accuracy() to compute common accuracy measures.
accuracy(car.lm.pred, valid.df$Price)
```
```{r}
car.lm.pred <- predict(car.lm, valid.df)
all.residuals <- valid.df$Price - car.lm.pred
hist(all.residuals, breaks = 25, xlab = "Residuals", main = "")
```

## Backward stepwise regression

Backward stepwise regression is a statistical approach employed in regression analysis to construct predictive models systematically by iteratively eliminating independent variables that contribute minimally to explaining the dependent variable's variance. This method begins with a full model encompassing all potential predictor variables and progressively removes the least influential ones. It finds utility in situations with multiple independent variables for several reasons: it simplifies complex models, enhances interpretability, guards against overfitting, and facilitates hypothesis testing by identifying the most significant variables in predicting the outcome of interest. In essence, backward stepwise regression streamlines models while preserving their predictive power, making it a valuable tool in statistical analysis.
```{r}
# use step() to run stepwise regression.
car.lm.step <- step(car.lm, direction = "backward")
summary(car.lm.step)  # Which variables were dropped?
```
```{r}
car.lm.step.pred <- predict(car.lm.step, valid.df)
accuracy(car.lm.step.pred, valid.df$Price)
```

## Regression model with no predictors 

Creating a regression model with no predictors, often called a null model, serves as a baseline for comparison in statistical analysis. It helps assess whether adding predictors significantly improves model performance, aids in hypothesis testing to determine if predictors have a meaningful impact, and simplifies the process of model selection by identifying which variables contribute meaningfully to explaining the dependent variable. Additionally, null models are valuable for control group analysis in experimental settings and for educational purposes to illustrate the concept of model complexity.

```{r}
car.lm.null <- lm(Price~1, data = train.df)
# use step() to run forward regression.
car.lm.step <- step(car.lm.null, scope=list(lower=car.lm.null, upper=car.lm), direction = "forward")
summary(car.lm.step)  # Which variables were added?
car.lm.step.pred <- predict(car.lm.step, valid.df)
accuracy(car.lm.step.pred, valid.df$Price)
```


```{r}
# use step() to run stepwise regression.
car.lm.step <- step(car.lm, direction = "both")
summary(car.lm.step)  # Which variables were dropped/added?
car.lm.step.pred <- predict(car.lm.step, valid.df)
accuracy(car.lm.step.pred, valid.df$Price)
```







